Linting Linearity in Core/System FC

### 2023-01-18 report.tex
Start writing about GADTs

Because Haskell is a pure programming language, input/ouptut side-effects are
"made pure" by modelling them on the type level through the
higher-kinded\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
constructor}} type \textbf{IO}. Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using Monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

The following function makes use of the above definitions to return the size of
a file given its path.
\begin{code}
countWords :: FilePath -> IO Integer
countWords path = do
    handle <- openFile path ReadMode
    size   <- hFileSize handle
    return size
\end{code}
% hClose handle

Another defining feature of Haskell is its powerful type system....

% TODO: File example

A function to compute the length of a list in Haskell is defined by pattern
matching on the list and returning $0$ if the list is empty and adding one to
the (computed recursively) length of the rest of the list otherwise. The
following snippet defines the said \textbf{length} function.
\begin{code}
    length [] = 0
    length (x:xs) = 1 + length xs
\end{code}
The function \textbf{length} receives a list and returns an integer, and since
it treats all list's elements the same, it is said to be polymorphic over the
list's elements type. In Haskell, function signatures typically appear above the
definition, and for this particular function we'd have
\begin{code}
    length :: [a] -> Int
\end{code}
where \textbf{a} is a universally quantified type variable, that is,
\textbf{length} can be called on lists of ints, lists of characters, lists of
lists of strings, lists of anything.

The laziness evaluation strategy entails expressions are only evaluated when
needed. If we were to call just the above function on a list of computationally
expensive elements, such as
\begin{code}
    length [5^1234, 6^4567, 8^(9^42)]
\end{code}
We would get the result 3 instantaneously: the elements of the list would never
be evaluated because the \textbf{length} function does not require the value of
each element to compute a result.

@@

\begin{itemize}
    \item Features fundamentais
    \item Exemplos, introduzir syntax
    \item (Ao contrário de OCaml) tem um sistema de tipos muito sufesticado
    \item Eventualmente explicação de GADTs? c/ exemplo do Vec
    \item Uma das features mais avaçadas/experimentais é o Linear Haskell
\end{itemize}

@@

\section{OutsideIn(X)}

>>>>

Because Haskell is a pure programming language, input/ouptut side-effects are
"made pure" by modelling them on the type level through the
higher-kinded\footnote{\textbf{IO} has kind $\textbf{Type}\to\textbf{Type}$,
that is, it is only a type after another type is passed as a parameter (e.g.
\textbf{IO~Int}, \textbf{IO~Bool}); \textbf{IO} by itself is a \emph{type
constructor}} type \textbf{IO}. Some of the example programs will look though as
if they had statements, but, in reality, the sequential appearence is just
syntatic sugar to an expression using monadic operators. The main take away is
that computations that do I/O may be sequenced together with other operations
that do I/O while retaining the lack of statements and the language purity
guarantees.

@@

The following function makes use of the above definitions to return the size of
a file given its path. Note that the function silently leaks the handle to the
file, despite compiling successfully.

\begin{code}
countWords :: FilePath -> IO Integer
countWords path = do
    handle <- openFile path ReadMode
    size   <- hFileSize handle
    return size
\end{code}
% hClose handle

Another defining feature of Haskell is its powerful type system. Features such
as type classes, arbitrary-rank types, functional dependencies, GADTs, type
families and (now) even impredicative polymorphism\cite{10.1145/3408971} have become
commonplace in Haskell code enforcing application level invariants and program
correctness through the types. For example, consider the definition of
\textbf{head} in the standard library, a function which takes the first element
of a list by pattern matching on the list constructors.

\begin{code}
    head :: [a] -> a
    head [] = error "List is empty!"
    head (x:xs) = x
\end{code}

If the list is empty, we terminate the program with an error, otherwise, we take
the first element. This function is unsafe -- our program might crash if we use
it on an invalid input. A common technique in Haskell is to use more expressive
types to assert properties about the values and get rid of the impossible cases
(e.g. a \textbf{NonEmpty} is a list that can not be empty). A well liked motto is
"make invalid states unrepresentable". In this light, we will introduce
Generalized Algebraic Data Types (GADTs) and create a list type indexed by size
for which we can write a completely safe \textbf{head} function by expressing
that the size of the list must be at least one, at the type level.

\textbf{GADTs}~\cite{10.1145/1159803.1159811,,,} are a wildly popular type level
feature that allow the user to define data types by enumerating its constructors
with explicit type signatures, where the result type may differ in the type
parameters across branches; pattern matching against GADTs might introduce type
refinements locally, i.e. in individual case alternatives. We develop the
length-indexed lists example without discussing the type system and type
inference details of GADTs which we explore later in~\ref{related-work-gadts}.
First, we define the natural numbers inductively: a natural number is either
zero (\textbf{Z}) or a successor (\textbf{S}) of another natural number (e.g.
the successor of the successor of zero is a natural number). The following
definition creates the \textbf{Nat} type, the \textbf{Z} and \textbf{S} data
constructors which construct values, \emph{and}, the \textbf{Z} and \textbf{S}
\emph{type} constructors which construct types. The latter two exist due to the
\emph{DataKinds} language feature which promotes term-level constructors to
type-level constructors.
%
\begin{code}
    data Nat = Z | S Nat
\end{code}
% Before introducing the data type of

Besides the aforementioned type-level features of Haskell, \emph{linear types},
which allow for a finer control over the usage of certain resources at the type
level, are another example of an extension to Haskell's type system that makes
it yet more expressive.

% A function to compute the length of a list in Haskell is defined by pattern
% matching on the list and returning $0$ if the list is empty and adding one to
% the (computed recursively) length of the rest of the list otherwise. The
% following snippet defines the said \textbf{length} function.
% \begin{code}
%     length [] = 0
%     length (x:xs) = 1 + length xs
% \end{code}
% The function \textbf{length} receives a list and returns an integer, and since
% it treats all list's elements the same, it is said to be polymorphic over the
% list's elements type. In Haskell, function signatures typically appear above the
% definition, and for this particular function we'd have
% \begin{code}
%     length :: [a] -> Int
% \end{code}
% where \textbf{a} is a universally quantified type variable, that is,
% \textbf{length} can be called on lists of ints, lists of characters, lists of
% lists of strings, lists of anything.

% The laziness evaluation strategy entails expressions are only evaluated when
% needed. If we were to call just the above function on a list of computationally
% expensive elements, such as
% \begin{code}
%     length [5^1234, 6^4567, 8^(9^42)]
% \end{code}
% We would get the result 3 instantaneously: the elements of the list would never
% be evaluated because the \textbf{length} function does not require the value of
% each element to compute a result.

@@

\begin{itemize}
    % \item Features fundamentais
    % \item Exemplos, introduzir syntax
    \item (Ao contrário de OCaml) tem um sistema de tipos muito sofisticado
    \item Eventualmente explicação de GADTs? c/ exemplo do Vec
    \item Uma das features mais avaçadas/experimentais é o Linear Haskell
\end{itemize}

@@

\section{OutsideIn(X)\label{related-work-gadts}}
